---
title: "Workshop"
author: "Lauren Portner"
date: "January 21, 2019"
output: html_document
---

## Creating objects in R

```{r arithmetic}


3 + 5
12 / 7

5 * 5

# assign value to object
weight_kg <- 55

2.2 * weight_kg

weight_lb <- 2.2 * weight_kg

sqrt(weight_kg)

round(3.14159)

round(3.14159, digits = 2)

round(3.14159, 2)
```

### Vectors and data types

This section will describe some basic data types in R

```{r}

weight_g <- c(50,60,65,82)

animals <- c("mouse","rat","dog")
```

Vector types in R:
* numeric
* character
* logical (TRUE or FALSE)
* Factors (categorical data)
* Dates

A vector is a data structure in R.

Other data structures:
* lists
* data frames
* matrices
* of course, vectors

Often, you want to convert lists and matrices to data frames or vectors.

## Data Frames

Nest, we're going to look at the structure of Data Frames.

```{r}
library(tidyverse)

download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv")

library(here) # This package makes working directories and file paths easy.

surveys <- read_csv(here("read_data", "portal_data_joined.csv"))

surveys

str (surveys)

dim(surveys)
nrow(surveys)
ncol(surveys)

summary(surveys)
```

### Indexing and subsetting data frames

First, let's use square bracket subsetting.

Square brackets are great for defining coordinates to extract data from

```{r setup warning = FALSE}

#First define the row coordinate, and then the column. Also write row and then column.
surveys[1,1]

surveys[1:3, 7]
# Give us all the rows and columns except column 7
surveys[,-7]
```


## Data Manipulation

Key functions for data manipulation

* select(): subsetting columns
* filter(): subsets of rows based on conditions
* mutate(): create new columns, based on information from other columns
* group_by(): creates groups based on catergorical data in a column
* summarize(): creates summary stats on grouped data
* arrange(): sort results
* count(): gives a count of discrete values

```{r}
select(surveys, plot_id, species_id, weight)

#Negative subsetting
select(surveys, -record_id)

filter(surveys, year == 1995)

```

## Pipes

Pipes allow you to chain together dplyr functions.

Pipe: %>% or ctrl-shift-m

```{r}
# Write multiple arguements in a sentence using pipes
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

Challenge # 1

Using pipes, subset the surveys dataframe to include animals collected in 1995 and retain only the columns year, sex and weight.

```{r challenge 1}
surveys %>%
  filter (year == 1995) %>%
  select (year, sex, weight)

surveys %>%
  mutate(weight_kg = weight / 1000, weight_kg2 = weight_kg * 2)

surveys %>%
  mutate(mean_weight = mean(weight)) %>%

surveys <- surveys %>%
  drop_na(weight) %>%
  mutate(mean_weight = mean(weight))
 
```

Challenge #2

Using the surveys data frame, create a new dataframe that contains only the species_id column,
has a new column called hindfoot_half: contains the values that are half the hindfoot_length values. Also, in the new hind_foot column there are no NAs and values are all less than 30

```{r}
surveys_hindfoot_half <- surveys %>%
  drop_na(hindfoot_length) %>%
  mutate(hindfoot_half = hindfoot_length / 2) %>%
  filter(hindfoot_half < 30) %>%
  select(species_id, hindfoot_half)

```

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))

surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>%
  arrange(min_weight)

```

```{r}
surveys %>%
  count (sex, sort = TRUE)

# the above code is synonymous with
surveys %>%
  group_by(sex) %>%
  summarize(count = n())
```

Challenge # 3

1. How many animals were caught in each plot type surveyed.

2. Use group_by and summarize to find mean, min and max of hindfoot_length (using species_id) for each species. Also, add the number of observations (hint: see ?n)

3. What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight.

```{r}
surveys %>%
  count(plot_type)

surveys %>%
  filter (!is.na(hindfoot_length)) %>%
  group_by (species_id) %>%
  summarize(
    mean_hindfoot_length = mean(hindfoot_length),
    min_hindfoot_length = min(hindfoot_length),
    max_hindfoot_length = max(hindfoot_length),
    n = n()
  )

max_weights <- surveys %>%
filter(!is.na(weight)) %>%
  group_by(year) %>%
  filter(weight == max(weight)) %>%
  select (year, genus, species, weight) %>%
  arrange (year)


```


## Export our data
```{r}
write_csv(max_weights, here("write_data","max_weights.csv"))
```




